---
title: "Homework Template"
output:
  html_document: default
  pdf_document: default
date: "2023-10-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
renv::restore()
#Installs all required packages
library(tidyverse)
library(tinytex)
library(dplyr)
library(janitor)
library(ggplot2)
library(here)
library(palmerpenguins)
library(ragg)
library(svglite)
source(here("functions", "cleaning.R"))
source(here("functions", "plotting.R"))
```

*The following is a template .rmd RMarkdown file for you to use for your homework submission.*

*Please Knit your .rmd to a PDF format or HTML and submit that with no identifiers like your name.*

*To create a PDF, first install tinytex and load the package. Then press the Knit arrow and select "Knit to PDF".*

## QUESTION 01: Data Visualisation for Science Communication

*Create a figure using the Palmer Penguin dataset that is correct but badly communicates the data. **Do not make a boxplot**.*

*Use the following references to guide you:*

-   [*https://www.nature.com/articles/533452a*](https://www.nature.com/articles/533452a){.uri}
-   [*https://elifesciences.org/articles/16800*](https://elifesciences.org/articles/16800){.uri}

*Note: Focus on visual elements rather than writing misleading text on it.*

### a) Provide your figure here:

```{r bad figure code, echo=FALSE}
#Loading the data (note: previous steps have been taken to ensure a clean csv)
penguins_clean <- read.csv(here("data", "penguins_clean.csv"))

bad_means <- aggregate(flipper_length_mm ~ species, data = penguins_clean, FUN = mean)

badplot <- ggplot(
  data = bad_means, 
  aes(x = species, 
      y = flipper_length_mm)) +
  geom_point()
badplot

# There is no need to provide the code for your bad figure, just use echo=FALSE so the code is hidden. Make sure your figure is visible after you knit it. 

```

### b) Write about how your design choices mislead the reader about the underlying data (200-300 words).

*There are three overarching mistakes made with the design choices for the above chart:*

-   Wrong chart type selection: A scatter plot is used, despite the fact that x-axis is categorical (species). Therefore the distance between dots does not show any useful information, and confuse the reader. Instead, a bar chart would be more appropriate.
-   Oversimplified data: By averaging the flipper lengths, the information shown is reduced to just a single point and therefore the variation, SD etc are all missing. Furthermore, the single dot does not convey the meaning of the data well, as a reader would struggle to realise that it represents the mean length. To correct this, a change in chart type as well as clear labeling is needed.
-   Truncated y-axis: The y-axis begins from 190, exaggerating the difference in flipper lengths between the different species. This misdirects the reader, as the actual difference between chinstrap and gentoo flipper length is 5% but the difference between dots is visually represented as being 200%.

*Sources:*

-   Sturge, Georgina. Bad Data. 3 Nov. 2022.

-   Keller, H., and Ch Trendelenburg. Data Presentation / Interpretation. Berlin, De Gruyter, 2019.

    ------------------------------------------------------------------------

## QUESTION 2: Data Pipeline

*Write a data analysis pipeline in your .rmd RMarkdown file. You should be aiming to write a clear explanation of the steps, the figures visible, as well as clear code.*

*Your code should include the steps practiced in the lab session:*

-   *Load the data*

-   *Appropriately clean the data*

-   *Create an Exploratory Figure (**not a boxplot**)*

-   *Save the figure*

-   ***New**: Run a statistical test*

-   ***New**: Create a Results Figure*

-   *Save the figure*

*An exploratory figure shows raw data, such as the distribution of the data. A results figure demonstrates the stats method chosen, and includes the results of the stats test.*

*Between your code, communicate clearly what you are doing and why.*

*Your text should include:*

-   *Introduction*

-   *Hypothesis*

-   *Stats Method*

-   *Results*

-   *Discussion*

-   *Conclusion*

*You will be marked on the following:*

### a) Your code for readability and functionality

### b) Your figures for communication

### c) Your text communication of your analysis

*Below is a template you can use.*

------------------------------------------------------------------------

### Introduction

```{r Data Exploration}

# Make sure your code prints. 

```

### Hypothesis

### Statistical Methods

```{r Statistics}

# Make sure your code prints. 

```

### Results & Discussion

```{r Plotting Results}

# Make sure your code prints. 

```

### Conclusion

------------------------------------------------------------------------

## QUESTION 3: Open Science

### a) GitHub

*Upload your RProject you created for **Question 2** and any files and subfolders used to GitHub. Do not include any identifiers such as your name. Make sure your GitHub repo is public.*

*GitHub link:*

*You will be marked on your repo organisation and readability.*

### b) Share your repo with a partner, download, and try to run their data pipeline.

*Partner's GitHub link:*

*You **must** provide this so I can verify there is no plagiarism between you and your partner.*

### c) Reflect on your experience running their code. (300-500 words)

-   *What elements of your partner's code helped you to understand their data pipeline?*

-   *Did it run? Did you need to fix anything?*

-   *What suggestions would you make for improving their code to make it more understandable or reproducible, and why?*

-   *If you needed to alter your partner's figure using their code, do you think that would be easy or difficult, and why?*

### d) Reflect on your own code based on your experience with your partner's code and their review of yours. (300-500 words)

-   *What improvements did they suggest, and do you agree?*

-   *What did you learn about writing code for other people?*
